ProteInfer, deep neural networks for protein functional inference

### 2023-02-28 latex/draft3.tex

\begin{figure}[h!]
  \centering
  \includegraphics[width=.4\columnwidth]{qr.pdf}

 \small QR code for the interactive version of this preprint at  \url{https://google-research.github.io/proteinfer/}
\end{figure}

@@

To address this challenge we employ deep dilated convolutional networks to learn the mapping between full-length protein sequences and functional annotations. The resulting ProteInfer models take amino acid sequences as input and are trained on the well-curated portion of the protein universe annotated by Swiss-Prot \cite{uniprotlatest}. We find that: 1) ProteInfer models reproduce curator decisions for a  variety of functional properties across sequences distant from the training data, 2) attribution analysis shows that the predictions are driven by relevant regions of each protein sequence, and 3) ProteInfer models create a generalised mapping between sequence space and the space of protein functions, which is useful for tasks other than those for which the models were trained. We provide trained ProteInfer networks that enable other researchers to reproduce the analysis presented and to explore embeddings of their proteins of interest, via both a command line tool\footnote{\url{https://github.com/google-research/proteinfer}}, and also via an in-browser JavaScript implementation that demonstrates the computational efficiency of deep-learning approaches.
\section*{Results}

\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}
%%%MODEL%%%
In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. In contrast, many previous approaches operate on fixed sequence lengths: these techniques are unable to make predictions for proteins larger than this sequence length, and use unnecessary resources when employed on smaller proteins. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities. 
We select all labels with predicted probability above a given confidence threshold, and varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

Each model was trained for about 60 hours using the Adam optimizer \citep{kingma2014adam} on 8 NVIDIA P100 GPUs with data parallelism \cite{dean2012large, dataparalellism}.  We found that using more than one GPU for training improved training time by allowing an increased batch size, but did not have a substantial impact on accuracy compared to training for longer with a smaller learning rate and smaller batch size on one GPU. The models have a small set of hyperparameters, such as the number of layers and the number of filters in each layer, which were tuned using random sampling to maximize $\texttt{F}_{max}$ on the random train-test split. Hyperparameter values are available in the \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
.
\subsection*{A machine-learning compatible dataset for protein function prediction}
The UniProt database is the central global repository for information about proteins. The manually curated portion, Swiss-Prot, is constructed by assessing $~$60,000 papers each year to harvest $\gt$ 35\% of the theoretically curatable information in the literature \cite{uniprotlatest}. We focus on Swiss-Prot to ensure that our models learn from human-curated labels, rather than labels generated by a computational annotation pipeline. Each protein in Swiss-Prot goes through a 6-stage process of sequence curation, sequence analysis, literature curation, family-based curation, evidence attribution, and quality assurance.
Functional annotation is stored in UniProt largely through \textit{database cross-references}, which link a specific protein with a label from a particular ontology. These cross-references include: Enzyme Commission (EC) numbers, representing the function of an enzyme; Gene Ontology (GO) terms relating to the protein's molecular function, biological process, or subcellular localisation; protein family information contained in the Pfam \citep{pfamlatest}, SUPFAM\citep{supfamlatest}, PRINTS \citep{prints}, TIGR \citep{tigr}, PANTHR \citep{panther} databases, or the umbrella database InterPro \citep{interpro}; as well as other information including ortholog databases and references on PubMed. Here, we focus on EC and GO labels, though our model training framework can immediately extend to other label sets.

@@

@@

Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

We implemented an alignment-based baseline in which BLASTp is used to identify the closest sequence to a query sequence in the train set. Labels are then imputed for the query sequence by transferring those labels that apply to the annotated match from the train set. We produced a precision-recall curve by using the bit score of the closest sequence as a measure of confidence, varying the cutoff above which we retain the imputed labels \cite{zhou2019cafa, eddy2011accelerated}. We also considered an ensemble of neural networks \cite{bileschi}, where the average of the ensemble elements' predicted probabilities is used as a confidence score, and a naive control where the number of proteins annotated with a specific term in the training set \citep{radivojac2013large}. %
%startleaveout
(See \Cref{fig:toplevel}B, \Cref{fig:with_ensemble}, \Cref{fig:without_ensemble}.)
%stopleaveout

%%%METHOD%%%
We found that BLASTp was able to achieve higher recall values than ProteInfer for lower precision values, while ProteInfer was able to provide greater precision than BLASTp at lower recall values. The high recall of BLAST is likely to reflect the fact that it has access to the entirety of the training set, rather than having to compress it into a limited set of neural network weights. In contrast, the lack of precision in BLAST could relate to reshuffling of sequences during evolution, which would allow a given protein to show high similarity to a trainining sequence in a particular subregion, despite lacking the core region required for that training sequence's function.  We wondered whether a combination of ProteInfer and BLASTp could synergize the best properties of both approaches. We found that even the simple ensembling strategy of rescaling the BLAST bit score by the averages of the ensembled CNNs' predicted probabilities gave a $\texttt{F}_{max}$ score (0.991, 95\% confidence interval [CI]: 0.990--0.992 ) that exceeded that of BLAST (0.984, 95\% CI: 0.983--0.985) or the ensembled CNN (0.981, 95\% CI: 0.980--0.982) alone (see Supplement for more details on this method). 
On the clustered train-test split based on UniRef50 (see \textit{clustered} in \Cref{fig:toplevel}B%htmlver:<a href="#fig5" target="_self">Fig. 5</a>
), we see a performance drop in all methods: this is expected, as remote homology tasks are designed to challenge methods to generalize farther in sequence space. The $\texttt{F}_{max}$ score of a single neural network fell to 0.914 (95\% CI: 0.913--0.915, precision: 0.959 recall: 0.875), substantially lower than BLAST (0.950, 95\% CI: 0.950--0.951), though again an ensemble of both BLAST and ProteInfer outperformed both (0.979, 95\% CI: 0.979--0.980). These patterns suggest that neural network methods learn different information about proteins to alignment-based methods, and so a combination of the two provides a synergistic result. All methods dramatically outperformed the naive frequency-based approach (\Cref{sup:fig:naive}%htmlver:see supplement
).

@@

We then assessed the ability of this method to more generally localize function within a sequence, even though the model was not trained with any explicit localization information. We selected all enzymes from Swiss-Prot that had two separate leaf-node EC labels for which our model predicted known EC labels, and these labels were mappable to corresponding Pfam labels. For each of these proteins, we obtained coarse-grained functional localization by using CAM to predict the order of the domains in the sequence and compared to the ground truth Pfam domain ordering\textnormal{(see supplement for details of the method)}%htmlver:(see <a href="./latex/supplement.pdf">supplement</a> for details of the method)
. We found that in 296 of 304 (97\%) of the cases, we correctly predicted the ordering, though we note that the set of bifunctional enzymes for which this analysis is applicable is limited in its functional diversity \textnormal{(see supplement)}%htmlver:(see <a href="./latex/supplement.pdf">supplement</a>)
. Although we did not find that fine-grained, per-residue functional localization arose from our application of CAM, we found that it reliably provided coarse-grained annotation of domains' order, as supported by Pfam. This experiment suggests that this is a promising future area for research.
\subsection*{Neural networks learn a general-purpose embedding space for protein function}
Whereas InterProScan compares each sequence against more than 50,000 %from summing integrated signatures at https://www.ebi.ac.uk/interpro/release_notes.html
individual signatures and BLAST compares against an even larger sequence database, ProteInfer uses a single deep model to extract features from sequences that directly predict protein function. One convenient property of this approach is that in the penultimate layer of the network each protein is expressed as a single point in a high-dimensional space. To investigate to what extent this space is useful in examining enzymatic function, we used the ProteInfer EC model trained on the random split to embed each test set protein sequence into a 1100-dimensional vector.

@@

%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} This figure shows EC-trained ProteInfer embeddings for all non-enzymatic sequences in the test set, projected using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, \emph{Q8RUD6} is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). \emph{Q8RUD6} can however act as a sulfertransferase \cite{pmid17408957}.

@@

The same trends for the relative performance of different approaches were seen for each of the direct-acyclic graphs that make up the GO ontology (biological process, cellular component and molecular function), but there were substantial differences in absolute performance (\Cref{sup:fig:go_ontologies}%htmlver:(see <a href="./latex/supplement.pdf">supplement</a>)
). Performance was highest for molecular function (max F1: 0.94), followed by biological process (max F1:0.86) and then cellular component (max F1:0.84).

@@

\begin{contributions}
  TS was responsible for conception, data ingestion, code, analysis, visualisation, UI, paper writing. MLB was
  responsible for visualization, analysis, code, CLI, open sourcing, testing, paper writing. DB was responsible for
  paper writing, analysis of results, code reviews. LJC was responsible for problem conception and framing, analysis of results, paper writing.
\end{contributions}

  \section*{Bibliography}
\bibliography{../bibliography}

\onecolumn
\newpage

\section*{Supplement}
\beginsupplement

\subsection{Input data statistics}

@@

\clearpage
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{ec_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{go_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{sequences_per_label_ec_random.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{sequences_per_label_go_random.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the random dataset.}per p
\end{figure}
\pagebreak
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{sequences_per_label_ec_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the clustered dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{sequences_per_label_go_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the clustered dataset.}
\end{figure}

When assigning examples to folds in our clustered dataset, we note that there are test examples that have labels that are never seen in the training data. We report these cases below as ``Impossible'' test example-label pairs.
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
                                Type &  Number \\
\Xhline{2pt}
                        Train labels &    3411 \\
\hline
                         Test labels &    3414 \\
\hline
              Impossible test labels &    1043 \\
\hline
           Train example-label pairs &  348105 \\
\hline
            Test example-label pairs &  348755 \\
\hline
 Impossible test example-label pairs &    3415 \\
\hline
\end{tabular}
\caption{Clustered dataset statistics for EC labels.}
\end{table}
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
                                Type &   Number \\
\Xhline{2pt}
                        Train labels &    26538 \\
\hline
                         Test labels &    26666 \\
\hline
              Impossible test labels &     3739 \\
\hline
           Train example-label pairs &  8338584 \\
\hline
            Test example-label pairs &  8424299 \\
\hline
 Impossible test example-label pairs &    11137 \\
\hline
\end{tabular}
\caption{Clustered dataset statistics for GO labels.}
\end{table}
\clearpage
\subsection{Precision/Recall curves}

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{random_truncated.pdf}
    \caption{EC number prediction: random split} 
    \label{fig:with_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{random_truncated_go.pdf}
    \caption{GO term prediction: random split} 
    \label{fig:with_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{clustered_truncated.pdf}
    \caption{EC number prediction: clustered split} 
    \label{fig:with_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{clustered_truncated_go.pdf}
    \caption{GO term prediction: clustered split} 
    \label{fig:with_ensemble:d} 
  \end{subfigure} 
  \caption{Bootstrapped precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs, and ensembled ProteInfer CNNs scaled by BLAST score. }
  \label{fig:with_ensemble} 
\end{figure}

>>>>

@@

To address this challenge we employ deep dilated convolutional networks to learn the mapping between full-length protein sequences and functional annotations. The resulting ProteInfer models take amino acid sequences as input and are trained on the well-curated portion of the protein universe annotated by Swiss-Prot \cite{uniprotlatest}. We find that: 1) ProteInfer models reproduce curator decisions for a  variety of functional properties across sequences distant from the training data, 2) attribution analysis shows that the predictions are driven by relevant regions of each protein sequence, and 3) ProteInfer models create a generalised mapping between sequence space and the space of protein functions, which is useful for tasks other than those for which the models were trained. We provide trained ProteInfer networks that enable other researchers to reproduce the analysis presented and to explore embeddings of their proteins of interest, via both a command line tool\footnote{\url{https://github.com/google-research/proteinfer}}, and also via an in-browser JavaScript implementation that demonstrates the computational efficiency of deep-learning approaches.
\section*{Methods}

\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}

In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. In contrast, many previous approaches operate on fixed sequence lengths: these techniques are unable to make predictions for proteins larger than this sequence length, and use unnecessary resources when employed on smaller proteins. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities.

%%%MODEL%%%

We select all labels with predicted probability above a given confidence threshold, and varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

Each model was trained for about 60 hours using the Adam optimizer \citep{kingma2014adam} on 8 NVIDIA P100 GPUs with data parallelism \cite{dean2012large, dataparalellism}.  We found that using more than one GPU for training improved training time by allowing an increased batch size, but did not have a substantial impact on accuracy compared to training for longer with a smaller learning rate and smaller batch size on one GPU. The models have a small set of hyperparameters, such as the number of layers and the number of filters in each layer, which were tuned using random sampling to maximize $\texttt{F}_{max}$ on the random train-test split. Hyperparameter values are available in the \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
.

\subsection*{A machine-learning compatible dataset for protein function prediction}
The UniProt database is the central global repository for information about proteins. The manually curated portion, Swiss-Prot, is constructed by assessing $~$60,000 papers each year to harvest $\gt$ 35\% of the theoretically curatable information in the literature \cite{uniprotlatest}. We focus on Swiss-Prot to ensure that our models learn from human-curated labels, rather than labels generated by a computational annotation pipeline. Each protein in Swiss-Prot goes through a 6-stage process of sequence curation, sequence analysis, literature curation, family-based curation, evidence attribution, and quality assurance.
Functional annotation is stored in UniProt largely through \textit{database cross-references}, which link a specific protein with a label from a particular ontology. These cross-references include: Enzyme Commission (EC) numbers, representing the function of an enzyme; Gene Ontology (GO) terms relating to the protein's molecular function, biological process, or subcellular localisation; protein family information contained in the Pfam \citep{pfamlatest}, SUPFAM\citep{supfamlatest}, PRINTS \citep{prints}, TIGR \citep{tigr}, PANTHR \citep{panther} databases, or the umbrella database InterPro \citep{interpro}; as well as other information including ortholog databases and references on PubMed. Here, we focus on EC and GO labels, though our model training framework can immediately extend to other label sets.

@@

\section*{Results}

@@

Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ methods}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

We implemented an alignment-based baseline in which BLASTp is used to identify the closest sequence to a query sequence in the train set. Labels are then imputed for the query sequence by transferring those labels that apply to the annotated match from the train set. We produced a precision-recall curve by using the bit score of the closest sequence as a measure of confidence, varying the cutoff above which we retain the imputed labels \cite{zhou2019cafa, eddy2011accelerated}. We also considered an ensemble of neural networks \cite{bileschi}, where the average of the ensemble elements' predicted probabilities is used as a confidence score, and a naive control where the number of proteins annotated with a specific term in the training set \citep{radivojac2013large}. %
%startleaveout
(See \Cref{fig:toplevel}B, \Cref{sup:fig:with_ensemble}, \Cref{sup:fig:without_ensemble}.)
%stopleaveout

%%%METHOD%%%
We found that BLASTp was able to achieve higher recall values than ProteInfer for lower precision values, while ProteInfer was able to provide greater precision than BLASTp at lower recall values. The high recall of BLAST is likely to reflect the fact that it has access to the entirety of the training set, rather than having to compress it into a limited set of neural network weights. In contrast, the lack of precision in BLAST could relate to reshuffling of sequences during evolution, which would allow a given protein to show high similarity to a trainining sequence in a particular subregion, despite lacking the core region required for that training sequence's function.  We wondered whether a combination of ProteInfer and BLASTp could synergize the best properties of both approaches. We found that even the simple ensembling strategy of rescaling the BLAST bit score by the averages of the ensembled CNNs' predicted probabilities gave a $\texttt{F}_{max}$ score (0.991, 95\% confidence interval [CI]: 0.990--0.992 ) that exceeded that of BLAST (0.984, 95\% CI: 0.983--0.985) or the ensembled CNN (0.981, 95\% CI: 0.980--0.982) alone (see Methods for more details on this method). 
On the clustered train-test split based on UniRef50 (see \textit{clustered} in \Cref{fig:toplevel}B%htmlver:<a href="#fig5" target="_self">Fig. 5</a>
), we see a performance drop in all methods: this is expected, as remote homology tasks are designed to challenge methods to generalize farther in sequence space. The $\texttt{F}_{max}$ score of a single neural network fell to 0.914 (95\% CI: 0.913--0.915, precision: 0.959 recall: 0.875), substantially lower than BLAST (0.950, 95\% CI: 0.950--0.951), though again an ensemble of both BLAST and ProteInfer outperformed both (0.979, 95\% CI: 0.979--0.980). These patterns suggest that neural network methods learn different information about proteins to alignment-based methods, and so a combination of the two provides a synergistic result. All methods dramatically outperformed the naive frequency-based approach (\Cref{sup:fig:naive}%htmlver:see supplement
).

@@

We then assessed the ability of this method to more generally localize function within a sequence, even though the model was not trained with any explicit localization information. We selected all enzymes from Swiss-Prot that had two separate leaf-node EC labels for which our model predicted known EC labels, and these labels were mappable to corresponding Pfam labels. For each of these proteins, we obtained coarse-grained functional localization by using CAM to predict the order of the domains in the sequence and compared to the ground truth Pfam domain ordering\textnormal{(see supplement for details of the method)} %htmlver:(see <a href="./latex/supplement.pdf">supplement</a> for details of the method)
. We found that in 296 of 304 (97\%) of the cases, we correctly predicted the ordering, though we note that the set of bifunctional enzymes for which this analysis is applicable is limited in its functional diversity \textnormal{(see methods)}%htmlver:(see <a href="./latex/supplement.pdf">supplement</a>)
. Although we did not find that fine-grained, per-residue functional localization arose from our application of CAM, we found that it reliably provided coarse-grained annotation of domains' order, as supported by Pfam. This experiment suggests that this is a promising future area for research.
\subsection*{Neural networks learn a general-purpose embedding space for protein function}
Whereas InterProScan compares each sequence against more than 50,000 %from summing integrated signatures at https://www.ebi.ac.uk/interpro/release_notes.html
individual signatures and BLAST compares against an even larger sequence database, ProteInfer uses a single deep model to extract features from sequences that directly predict protein function. One convenient property of this approach is that in the penultimate layer of the network each protein is expressed as a single point in a high-dimensional space. To investigate to what extent this space is useful in examining enzymatic function, we used the ProteInfer EC model trained on the random split to embed each test set protein sequence into a 1100-dimensional vector.

@@

%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \includegraphics[width=0.9\textwidth]{combined_subfigs_a.pdf}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} This figure shows EC-trained ProteInfer embeddings for all non-enzymatic sequences in the test set, projected using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, \emph{Q8RUD6} is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). \emph{Q8RUD6} can however act as a sulfertransferase \cite{pmid17408957}.

@@

The same trends for the relative performance of different approaches were seen for each of the direct-acyclic graphs that make up the GO ontology (biological process, cellular component and molecular function), but there were substantial differences in absolute performance (\Cref{sup:fig:go_ontologies}%htmlver: (see <a href="./latex/supplement.pdf">supplement</a>)
). Performance was highest for molecular function (max F1: 0.94), followed by biological process (max F1:0.86) and then cellular component (max F1:0.84).

@@

\subsection*{Input data statistics}

@@

  \begin{table}[htbp]
    \centering
    \begin{tabular}{|l|r|}
    \hline
                                    Type &  Number \\
    \Xhline{2pt}
                            Train labels &    3411 \\
    \hline
                             Test labels &    3414 \\
    \hline
                  Impossible test labels &    1043 \\
    \hline
               Train example-label pairs &  348105 \\
    \hline
                Test example-label pairs &  348755 \\
    \hline
     Impossible test example-label pairs &    3415 \\
    \hline
    \end{tabular}
    \caption{Clustered dataset statistics for EC labels.}
    \end{table}
  \begin{table}[htbp]
  \centering
  \begin{tabular}{|l|r|}
  \hline
                                  Type &   Number \\
  \Xhline{2pt}
                          Train labels &    26538 \\
  \hline
                           Test labels &    26666 \\
  \hline
                Impossible test labels &     3739 \\
  \hline
             Train example-label pairs &  8338584 \\
  \hline
              Test example-label pairs &  8424299 \\
  \hline
   Impossible test example-label pairs &    11137 \\
  \hline
  \end{tabular}
  \caption{Clustered dataset statistics for GO labels.}
  \end{table}

  \begin{table}[htbp]
    \centering
  \begin{tabular}{|l|r|}
  \hline
  Vocabulary & Number of terms \\
  \hline
  EC & 5,134 \\
  \hline
  GO & 32,109 \\
  \hline
  \end{tabular}
  \caption{Vocabulary sizes in models trained for EC and GO.}
  \label{sup:table:vocab_sizes}
  \end{table}

When assigning examples to folds in our clustered dataset, we note that there are test examples that have labels that are never seen in the training data. We report these cases below as ``Impossible'' test example-label pairs.
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
                                Type &  Number \\
\Xhline{2pt}
                        Train labels &    3411 \\
\hline
                         Test labels &    3414 \\
\hline
              Impossible test labels &    1043 \\
\hline
           Train example-label pairs &  348105 \\
\hline
            Test example-label pairs &  348755 \\
\hline
 Impossible test example-label pairs &    3415 \\
\hline
\end{tabular}
\end{table}

\subsection*{Domain architecture diversity in bifunctional enzymes}
\begin{table}[htbp]
  \begin{tabular}{|p{0.2\linewidth}|p{0.2\linewidth}|p{0.12\linewidth}|p{0.12\linewidth}|p{0.12\linewidth}|}
  \hline
  First domain & Second Domain & Number ordered correctly & Number times seen & Percent Correct \\
  \hline
  EC:2.7.7.60  & EC:4.6.1.12   & 94                                           & 94                                       & 100                                 \\
  \hline
  EC:4.1.99.12 & EC:3.5.4.25   & 83                                           & 83                                       & 100                                 \\
  \hline
  EC:3.5.4.19  & EC:3.6.1.31   & 59                                           & 59                                       & 100                                 \\
  \hline
  EC:1.8.4.11  & EC:1.8.4.12   & 20                                           & 20                                       & 100                                 \\
  \hline
  EC:4.1.1.48  & EC:5.3.1.24   & 18                                           & 18                                       & 100                                 \\
  \hline
  EC:5.4.99.5  & EC:4.2.1.51   & 12                                           & 12                                       & 100                                 \\
  \hline
  EC:5.4.99.5  & EC:1.3.1.12   & 4                                            & 4                                        & 100                                 \\
  \hline
  EC:4.2.1.10  & EC:1.1.1.25   & 3                                            & 3                                        & 100                                 \\
  \hline
  EC:2.7.7.61  & EC:2.4.2.52   & 0                                            & 3                                        & 0                                   \\
  \hline
  EC:2.7.1.71  & EC:4.2.3.4    & 0                                            & 2                                        & 0                                   \\
  \hline
  EC:1.1.1.25  & EC:4.2.1.10   & 0                                            & 1                                        & 0                                   \\
  \hline
  EC:2.7.2.3   & EC:5.3.1.1    & 1                                            & 1                                        & 100                                 \\
  \hline
  EC:4.1.1.97  & EC:1.7.3.3    & 1                                            & 1                                        & 100                                 \\
  \hline
  EC:4.1.3.1   & EC:2.3.3.9    & 1                                            & 1                                        & 100                                 \\
  \hline
  EC:5.1.99.6  & EC:1.4.3.5    & 0                                            & 1                                        & 0                                  \\
  \hline
  EC:1.8.4.12  & EC:1.8.4.11   & 0                                            & 1                                        & 0                                   \\
  \hline
  \end{tabular}
  \caption{Domain architecture diversity in bifunctional enzymes. In Swiss-Prot, there are 16 candidate domain architectures available for our EC functional localization experiment. Among these all domain architectures with more than 3 instances in Swiss-Prot (7 of them) are 100\% correctly ordered by our CAM method.}
  \end{table}

\subsection*{Combining CNN and BLAST}
We noticed a clear difference between the performance of CNN models and BLAST models. The CNN models were able make predictions with higher precision than BLAST for a given recall value, but achieved lower overall recall, while BLAST gave high recall at the expense of precision. This suggested that combining the two methods would be beneficial. A CNN, or an ensemble of CNNs produces a metric that is notionally a probability (though often imperfectly calibrated), while BLAST bit-score produces a bit-score metric indicating the significance of the match. We reasoned that one approach to combining the two would simply be to multiply the values together - improving the precision of BLAST predictions by reducing bit-scores in cases where the CNN model lacked confidence in a prediction. This approach performed well, and was the best that we evaluated. To implement this method, all labels associated with the top hit from BLAST with a sequence are initially assigned identical scores, determined by the bit-score of the top match, but these are then rescaled according to the output of the ProteInfer command-line interface.
